---
# =============================================================================
# Script:      report_template.Rmd
# Pipeline:    16S Microbiota Pipeline
# Author:      Roberto C. Torres, PhD. <torres.roberto.c@gmail.com>
# Institution: Bioinformatics Lab, Infectious Diseases Research Unit,
#              CMN SXXI, IMSS, Mexico City
# Description: flexdashboard template for interactive HTML report of pipeline results
# =============================================================================
title: "16S Microbiota Analysis Pipeline — Results Report"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    theme: flatly
    self_contained: true
    navbar:
      - { title: "Summary", href: "#resumen" }
      - { title: "Medical Research Unit on Infectious and Parasitic Diseases (UIMEIP) · IMSS · Mexico City", align: right }
params:
  out_dir: ""
  case_label: "Case"
  ctrl_label: "Control"
  group_col: "Group"
  meta_file: ""
  sample_id_col: "Sample_ID"
  sample_type_filter: ~
  sample_type_col: "sample_type"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

load_pkg <- function(pkg) {
  ok <- requireNamespace(pkg, quietly = TRUE)
  if (ok) suppressPackageStartupMessages(library(pkg, character.only = TRUE))
  ok
}

pkg_status <- c(
  flexdashboard = load_pkg("flexdashboard"),
  ggplot2 = load_pkg("ggplot2"),
  plotly = load_pkg("plotly"),
  DT = load_pkg("DT"),
  dplyr = load_pkg("dplyr"),
  tidyr = load_pkg("tidyr"),
  knitr = load_pkg("knitr"),
  base64enc = load_pkg("base64enc")
)

f <- function(...) file.path(params$out_dir, ...)
safe_read <- function(path, ...) {
  if (!file.exists(path)) return(NULL)
  tryCatch(
    read.table(
      path,
      header = TRUE,
      stringsAsFactors = FALSE,
      check.names = FALSE,
      ...
    ),
    error = function(e) NULL
  )
}

safe_read_csv <- function(path, ...) safe_read(path, sep = ",", ...)

group_levels <- c(params$ctrl_label, params$case_label)
distinct_colors <- function(n, seed = 42L) {
  if (n <= 0L) return(character(0))
  if (requireNamespace("randomcoloR", quietly = TRUE)) {
    set.seed(seed)
    randomcoloR::distinctColorPalette(n)
  } else {
    base_cols <- grDevices::colors()
    base_cols <- base_cols[!grepl("white|grey|gray|light|pale", base_cols)]
    set.seed(seed)
    sample(base_cols, min(n, length(base_cols)), replace = n > length(base_cols))
  }
}
run_seed <- suppressWarnings(as.integer(gsub("[^0-9]", "", basename(params$out_dir))))
set.seed(if (!is.na(run_seed) && run_seed > 0L) run_seed else 42L)
group_colors <- setNames(distinct_colors(2L, seed = if (!is.na(run_seed) && run_seed > 0L) run_seed else 42L), group_levels)

has_pkgs <- function(pkgs) all(vapply(pkgs, function(x) isTRUE(pkg_status[[x]]) || requireNamespace(x, quietly = TRUE), logical(1)))

box_tag <- function(text, type = "warning", title = NULL) {
  if (requireNamespace("htmltools", quietly = TRUE)) {
    bg <- switch(type, error = "#f8d7da", warning = "#fff3cd", info = "#d1ecf1", ok = "#d4edda", "#fff3cd")
    border <- switch(type, error = "#f5c6cb", warning = "#ffeeba", info = "#bee5eb", ok = "#c3e6cb", "#ffeeba")
    htmltools::div(
      style = paste(
        "padding:12px;border-radius:6px;border:1px solid", border,
        ";background:", bg, ";margin:8px 0;"
      ),
      if (!is.null(title)) htmltools::tags$strong(paste0(title, " ")),
      text
    )
  } else {
    paste(if (!is.null(title)) paste0(title, ": "), text)
  }
}

fmt_n <- function(x, digits = 3) {
  if (length(x) == 0 || all(is.na(x))) return("NA")
  if (is.numeric(x)) return(format(round(x[1], digits), trim = TRUE, scientific = FALSE))
  as.character(x[1])
}

pct_lab <- function(x) paste0(round(100 * x), "%")

flags_val <- function(key, step = NULL, default = NA_character_) {
  if (is.null(flags_all) || !nrow(flags_all)) return(default)
  idx <- flags_all$key == key
  if (!is.null(step)) idx <- idx & flags_all$step == step
  vals <- flags_all$value[idx]
  if (!length(vals)) default else vals[1]
}

meta <- if (nzchar(params$meta_file)) safe_read_csv(params$meta_file) else NULL
tss <- safe_read(f("filter", "asv_table_tss.tsv"), sep = "\t")
tax <- safe_read(f("filter", "taxonomy_filtered.tsv"), sep = "\t")
alpha_div <- safe_read(f("alpha", "alpha_diversity.tsv"), sep = "\t")
alpha_wil <- safe_read(f("alpha", "alpha_wilcoxon.tsv"), sep = "\t")
permanova <- safe_read(f("beta", "permanova.tsv"), sep = "\t")
betadisper <- safe_read(f("beta", "betadisper.tsv"), sep = "\t")
pcoa_wunifrac <- safe_read(f("beta", "pcoa_wunifrac.tsv"), sep = "\t")
pcoa_uwunifrac <- safe_read(f("beta", "pcoa_uwunifrac.tsv"), sep = "\t")
info_pcoa <- safe_read(f("informative", "informative_pcoa.tsv"), sep = "\t")
prev_tbl <- safe_read(f("composition", "species_prevalence_mean_relative_abundance.tsv"), sep = "\t")
top20_mean <- safe_read(f("composition", "top20_species_group_mean.tsv"), sep = "\t")
case_unique <- safe_read(f("composition", "species_unique_cases.tsv"), sep = "\t")
ctrl_unique <- safe_read(f("composition", "species_unique_controls.tsv"), sep = "\t")
da_res <- safe_read(f("differential", "da_results.tsv"), sep = "\t")
rf_imp <- safe_read(f("informative", "rf_importance.tsv"), sep = "\t")
lda_scores <- safe_read(f("informative", "lda_scores.tsv"), sep = "\t")
rf_means <- safe_read(f("informative", "rf_top_features_group_mean.tsv"), sep = "\t")

flag_files <- list.files(f("flags"), pattern = "^flags_.*\\.tsv$", full.names = TRUE)
flags_all <- if (length(flag_files)) {
  do.call(
    rbind,
    Filter(Negate(is.null), lapply(flag_files, function(fp) safe_read(fp, sep = "\t")))
  )
} else {
  NULL
}

if (!is.null(meta) && params$group_col %in% colnames(meta)) {
  meta <- meta[meta[[params$group_col]] %in% group_levels, , drop = FALSE]
  if (!is.null(params$sample_type_filter) && length(params$sample_type_filter) > 0 &&
      nzchar(params$sample_type_filter[[1]]) &&
      params$sample_type_col %in% colnames(meta)) {
    meta <- meta[meta[[params$sample_type_col]] %in% params$sample_type_filter, , drop = FALSE]
  }
}

# Contar solo muestras presentes en la tabla TSS (ground truth)
if (!is.null(tss) && "sample_id" %in% colnames(tss) && !is.null(meta) &&
    params$sample_id_col %in% colnames(meta) && params$group_col %in% colnames(meta)) {
  analyzed_meta <- meta[meta[[params$sample_id_col]] %in% tss$sample_id, , drop = FALSE]
  n_case <- sum(analyzed_meta[[params$group_col]] == params$case_label, na.rm = TRUE)
  n_ctrl <- sum(analyzed_meta[[params$group_col]] == params$ctrl_label, na.rm = TRUE)
} else if (!is.null(meta) && params$group_col %in% colnames(meta)) {
  n_case <- sum(meta[[params$group_col]] == params$case_label, na.rm = TRUE)
  n_ctrl <- sum(meta[[params$group_col]] == params$ctrl_label, na.rm = TRUE)
} else {
  n_case <- NA_integer_
  n_ctrl <- NA_integer_
}

n_species_retained <- if (!is.null(tss)) {
  max(0L, ncol(tss) - 1L)
} else if (!is.null(tax)) {
  nrow(tax)
} else {
  NA_integer_
}

build_dt <- function(df, caption = NULL, color_status = FALSE) {
  if (!isTRUE(pkg_status[["DT"]]) || is.null(df) || !nrow(df)) {
    return(box_tag("Table not available for this section.", "warning"))
  }
  tab <- DT::datatable(
    df,
    rownames = FALSE,
    caption = caption,
    options = list(pageLength = 8, scrollX = TRUE, dom = "tip")
  )
  if (color_status && "status" %in% colnames(df)) {
    tab <- DT::formatStyle(
      tab,
      "status",
      backgroundColor = DT::styleEqual(
        c("ok", "warn", "error"),
        c(NA, "#fff3cd", "#f8d7da")
      )
    )
  }
  tab
}

make_value_box <- function(value, caption, color) {
  if (isTRUE(pkg_status[["flexdashboard"]])) {
    flexdashboard::valueBox(value = value, caption = caption, color = color)
  } else {
    box_tag(paste(caption, value), type = "info")
  }
}
```

Summary {#resumen data-navmenu="Summary" data-icon="fa-home"}
=====================================

Column {data-width=280}
-------------------------------------

### About this report

This report summarizes the results of the 16S microbiota diversity analysis
pipeline. Each section corresponds to one analysis module.

**Groups:**

- **`r params$case_label`**: N = `r fmt_n(n_case, 0)`
- **`r params$ctrl_label`**: N = `r fmt_n(n_ctrl, 0)`

**Results directory:**  
`r basename(params$out_dir)`

### Contact

**Roberto C. Torres, PhD.**  
Medical Research Unit on Infectious and Parasitic Diseases (UIMEIP)  
Instituto Mexicano del Seguro Social (IMSS)  
Mexico City, Mexico  
[torres.roberto.c@gmail.com](mailto:torres.roberto.c@gmail.com)

Column {data-width=720}
-------------------------------------

### Analysis Parameters

```{r resumen-analisis}
# Leer region y kit desde flags
region_val  <- flags_val("region",  default = flags_val("REGION",  default = "—"))
kit_val     <- flags_val("kit",     default = flags_val("KIT",     default = "—"))
trunc_r1    <- flags_val("trunc_r1", step = "01_qc", default = "—")
trunc_r2    <- flags_val("trunc_r2", step = "01_qc", default = "—")
n_qc        <- flags_val("n_qc",    step = "01_qc", default = "—")
stype_flag  <- flags_val("sample_type_filter", step = "01_qc", default = "all")
n_raw       <- flags_val("n_samples_raw", step = "01_qc", default = "—")
n_filt      <- flags_val("n_samples_after_filter", step = "01_qc", default = "—")

items <- list(
  c("Analyzed sample type",         stype_flag),
  c(paste0(params$case_label, " samples (in table)"), fmt_n(n_case, 0)),
  c(paste0(params$ctrl_label, " samples (in table)"), fmt_n(n_ctrl, 0)),
  c("Total samples (manifest)",     n_raw),
  c("Samples after filtering",      n_filt),
  c("Hypervariable region",         region_val),
  c("Sequencing kit",               kit_val),
  c("TRUNC_LEN R1 / R2",           paste(trunc_r1, "/", trunc_r2)),
  c("Retained species/features",    fmt_n(n_species_retained, 0)),
  c("Results directory",            basename(params$out_dir))
)

df_sum <- as.data.frame(do.call(rbind, items), stringsAsFactors = FALSE)
colnames(df_sum) <- c("Parameter", "Value")

if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(df_sum, format = "html", escape = FALSE) |>
    (\(x) if (requireNamespace("kableExtra", quietly = TRUE))
      kableExtra::kable_styling(x, bootstrap_options = c("striped","hover"), full_width = FALSE)
    else x)()
} else {
  build_dt(df_sum)
}
```

Quality Control {#qc data-navmenu="QC" data-icon="fa-chart-line"}
=====================================

Column {data-width=280}
-------------------------------------

### Quality Control

```{r qc-desc}
trunc_r1_v <- flags_val("trunc_r1", step = "01_qc", default = "N/A")
trunc_r2_v <- flags_val("trunc_r2", step = "01_qc", default = "N/A")

if (requireNamespace("htmltools", quietly = TRUE)) {
  htmltools::tagList(
    htmltools::HTML(paste0(
      "<p><strong>Fig. 1.</strong> Quality Phred profiles (median ± IQR) for R1 and R2 reads, calculated ",
      "on a representative subsample. The dashed red line marks the quality threshold (Q25); ",
      "the dashed green line shows the suggested truncation point. Lower panels show read ",
      "length distributions. The orange dashed line marks the 5th percentile (p05).</p>",
      "<p><strong>Suggested truncation:</strong> R1 = ", trunc_r1_v, " bp / R2 = ", trunc_r2_v, " bp</p>"
    ))
  )
} else {
  cat(paste("Suggested truncation: R1 =", trunc_r1_v, "bp / R2 =", trunc_r2_v, "bp"))
}
```

Column {data-width=720}
-------------------------------------

### Fig. 1. Sequencing Quality Control
```{r qc-embed, fig.height=9, fig.width=13}
qc_pdf <- f("qc", "qc_report.pdf")
if (file.exists(qc_pdf)) {
  if (requireNamespace("base64enc", quietly = TRUE) && requireNamespace("htmltools", quietly = TRUE)) {
    b64 <- base64enc::base64encode(qc_pdf)
    htmltools::tags$iframe(
      src = paste0("data:application/pdf;base64,", b64),
      width  = "100%",
      height = "900px",
      style  = "border:none;"
    )
  } else {
    box_tag("The QC PDF exists, but it could not be embedded because base64enc and/or htmltools are missing.", "warning")
  }
} else {
  box_tag(paste0(
    "qc/qc_report.pdf was not found in the results directory. ",
    "Please verify that step 01_qc.R completed successfully."
  ), "warning")
}
```

Taxonomic Composition {data-navmenu="Composition"}
=====================================

Column {data-width=280}
-------------------------------------

### Taxonomic Composition

**Fig. 1.** Venn diagram showing species overlap between groups.

**Fig. 2.** Stacked bar chart of mean relative abundance for the top N species
per group (TSS-normalized).

**Fig. 3.** Scatter plot comparing per-species prevalence between groups. Species
detected exclusively in one group appear on the axes.

Column {data-width=720}
-------------------------------------

### Fig. 1. Venn Diagram of Species Overlap Between Groups

```{r venn-especies}
req_venn <- c("ggplot2")
if (is.null(prev_tbl) || !nrow(prev_tbl)) {
  box_tag("No prevalence table is available to generate the Venn diagram.", "warning")
} else {
  case_sp <- unique(prev_tbl$taxon[
    prev_tbl[[params$group_col]] == params$case_label & prev_tbl$prevalence > 0
  ])
  ctrl_sp <- unique(prev_tbl$taxon[
    prev_tbl[[params$group_col]] == params$ctrl_label & prev_tbl$prevalence > 0
  ])
  n_shared   <- length(intersect(case_sp, ctrl_sp))
  n_case_exc <- length(setdiff(case_sp, ctrl_sp))
  n_ctrl_exc <- length(setdiff(ctrl_sp, case_sp))

  if (requireNamespace("ggvenn", quietly = TRUE)) {
    venn_fill <- distinct_colors(2L, seed = if (!is.na(run_seed) && run_seed > 0L) run_seed else 42L)
    venn_list <- setNames(list(case_sp, ctrl_sp), c(params$case_label, params$ctrl_label))
    p_venn <- ggvenn::ggvenn(
      venn_list,
      fill_color   = unname(venn_fill),
      stroke_size  = 0.5,
      set_name_size = 4
    ) +
      ggplot2::labs(
        caption = paste0(
          "Shared: ", n_shared,
          " | Exclusive to ", params$case_label, ": ", n_case_exc,
          " | Exclusive to ", params$ctrl_label, ": ", n_ctrl_exc
        )
      ) +
      ggplot2::theme(plot.caption = ggplot2::element_text(size = 9))
    print(p_venn)
  } else {
    df_venn <- data.frame(
      Category = c(
        paste0("Exclusive to ", params$case_label),
        "Shared",
        paste0("Exclusive to ", params$ctrl_label)
      ),
      N = c(n_case_exc, n_shared, n_ctrl_exc)
    )
    build_dt(df_venn, caption = "Species overlap between groups")
  }
}
```

### Fig. 2. Top Species by Mean Relative Abundance

```{r top-especies-abundancia}
req <- c("ggplot2", "dplyr")
if (!has_pkgs(req) || is.null(top20_mean) || !nrow(top20_mean)) {
  box_tag("There are not enough data to reconstruct the top taxonomic composition.", "warning")
} else {
  taxa <- unique(as.character(top20_mean$taxon))
  taxa_palette <- setNames(distinct_colors(length(taxa), seed = if (!is.na(run_seed) && run_seed > 0L) run_seed else 42L), taxa)
  p_top20 <- ggplot2::ggplot(
    top20_mean,
    ggplot2::aes(
      x = .data[[params$group_col]],
      y = mean_rel_abundance,
      fill = taxon,
      text = paste0(
        "Group: ", .data[[params$group_col]],
        "<br>Taxon: ", taxon,
        "<br>Mean abundance: ", signif(mean_rel_abundance, 3)
      )
    )
  ) +
    ggplot2::geom_bar(stat = "identity", position = "fill", width = 0.8) +
    ggplot2::scale_fill_manual(values = taxa_palette) +
    ggplot2::scale_y_continuous(labels = pct_lab) +
    ggplot2::labs(
      title = paste0("Top ", length(taxa), " species by mean relative abundance"),
      x = "",
      y = "Fraction of total",
      fill = NULL
    ) +
    ggplot2::theme_minimal(base_size = 10)
  if (isTRUE(pkg_status[["plotly"]])) {
    plotly::ggplotly(p_top20, tooltip = "text")
  } else {
    p_top20
  }
}
```

### Fig. 3. Prevalence Scatter Plot Between Groups

```{r dispersion-prevalencia-grupos}
req <- c("ggplot2", "dplyr", "tidyr")
if (!has_pkgs(req) || is.null(prev_tbl) || !nrow(prev_tbl)) {
  box_tag("No prevalence table is available for this section.", "warning")
} else {
  prev_wide <- tidyr::pivot_wider(
    prev_tbl[, intersect(c("taxon", params$group_col, "prevalence", "class"), colnames(prev_tbl)), drop = FALSE],
    names_from = dplyr::all_of(params$group_col),
    values_from = prevalence,
    values_fill = 0
  )
  if (!all(c(params$ctrl_label, params$case_label) %in% colnames(prev_wide))) {
    box_tag("The prevalence table does not contain both expected group columns.", "warning")
  } else {
    prev_wide$ctrl_prev <- prev_wide[[params$ctrl_label]]
    prev_wide$case_prev <- prev_wide[[params$case_label]]
    prev_wide$class <- if ("class" %in% colnames(prev_wide)) prev_wide$class else "Shared"
    label_df <- prev_wide[prev_wide$case_prev > 10 | prev_wide$ctrl_prev > 10, , drop = FALSE]
    cls_cols <- c(Shared = "grey60")
    cls_cols[params$case_label] <- unname(group_colors[params$case_label])
    cls_cols[params$ctrl_label] <- unname(group_colors[params$ctrl_label])
    p_prev <- ggplot2::ggplot(
      prev_wide,
      ggplot2::aes(
        x = ctrl_prev,
        y = case_prev,
        color = class,
        text = paste0(
          "Taxon: ", taxon,
          "<br>", params$ctrl_label, ": ", ctrl_prev,
          "<br>", params$case_label, ": ", case_prev,
          "<br>Class: ", class
        )
      )
    ) +
      ggplot2::geom_point(size = 2.2, alpha = 0.8) +
      ggplot2::scale_color_manual(values = cls_cols, drop = FALSE) +
      ggplot2::labs(
        title = "Per-species prevalence between groups",
        x = paste0("Prevalence in ", params$ctrl_label),
        y = paste0("Prevalence in ", params$case_label),
        color = "Class"
      ) +
      ggplot2::theme_minimal(base_size = 10)
    if (requireNamespace("ggrepel", quietly = TRUE) && nrow(label_df)) {
      p_prev <- p_prev + ggrepel::geom_text_repel(
        data = label_df,
        ggplot2::aes(label = taxon),
        size = 2.8,
        show.legend = FALSE
      )
    }
    if (isTRUE(pkg_status[["plotly"]])) {
      plotly::ggplotly(p_prev, tooltip = "text")
    } else {
      p_prev
    }
  }
}
```

Alpha Diversity {data-navmenu="Diversity"}
=====================================

Column {data-width=280}
-------------------------------------

### Alpha Diversity

Violin plots show the distribution of four alpha diversity indices per group.
The inner box shows median and IQR; jittered points represent individual samples.
P-value annotations correspond to Wilcoxon tests with Benjamini-Hochberg correction.

**Fig. 5.** Alpha diversity metrics: Observed richness, Pielou evenness, Shannon H',
and Simpson 1-D.

Column {data-width=720}
-------------------------------------

### Alpha Diversity Metrics

```{r alpha-metricas}
req <- c("ggplot2")
if (!has_pkgs(req) || is.null(alpha_div) || !nrow(alpha_div) || is.null(alpha_wil) || !nrow(alpha_wil)) {
  box_tag("Complete alpha diversity data are not available to build all four plots.", "warning")
} else {
  metrics <- c(observed = "Observed richness", evenness = "Pielou evenness", shannon = "Shannon H'", simpson = "Simpson 1-D")
  alpha_div <- alpha_div[alpha_div[[params$group_col]] %in% group_levels, , drop = FALSE]
  alpha_div[[params$group_col]] <- factor(alpha_div[[params$group_col]], levels = group_levels)
  plot_metric <- function(metric, label) {
    p_adj_val <- alpha_wil$p_adj[alpha_wil$metric == metric]
    p_adj_val <- if (length(p_adj_val)) p_adj_val[1] else NA_real_
    p_label   <- if (!is.na(p_adj_val)) {
      if      (p_adj_val < 0.001) "p < 0.001"
      else if (p_adj_val < 0.05)  paste0("p = ", signif(p_adj_val, 2))
      else                         paste0("p = ", signif(p_adj_val, 2), " (ns)")
    } else { "p = N/A" }
    y_max <- max(alpha_div[[metric]], na.rm = TRUE)
    y_ann <- y_max * 1.08
    ggplot2::ggplot(
      alpha_div,
      ggplot2::aes(
        x = .data[[params$group_col]],
        y = .data[[metric]],
        fill = .data[[params$group_col]],
        text = paste0(
          "Group: ", .data[[params$group_col]],
          "<br>", label, ": ", signif(.data[[metric]], 4)
        )
      )
    ) +
      ggplot2::geom_violin(trim = FALSE, alpha = 0.55) +
      ggplot2::geom_boxplot(width = 0.14, outlier.shape = NA, alpha = 0.85, color = "black") +
      ggplot2::geom_jitter(width = 0.1, size = 1, alpha = 0.4, show.legend = FALSE) +
      ggplot2::annotate("segment",
        x = 1, xend = 2, y = y_ann, yend = y_ann, color = "black") +
      ggplot2::annotate("text",
        x = 1.5, y = y_ann * 1.02, label = p_label, size = 3.2) +
      ggplot2::scale_fill_manual(values = group_colors) +
      ggplot2::labs(title = label, x = "", y = label) +
      ggplot2::theme_minimal(base_size = 10) +
      ggplot2::theme(legend.position = "none")
  }
  p_list <- Map(plot_metric, names(metrics), unname(metrics))
  if (isTRUE(pkg_status[["plotly"]])) {
    # geom_violin fails in ggplotly with small groups; use boxplot+jitter for interactive version
    plot_metric_plotly <- function(metric, label) {
      p_adj_val <- alpha_wil$p_adj[alpha_wil$metric == metric]
      p_adj_val <- if (length(p_adj_val)) p_adj_val[1] else NA_real_
      p_label   <- if (!is.na(p_adj_val)) {
        if      (p_adj_val < 0.001) "p < 0.001"
        else if (p_adj_val < 0.05)  paste0("p = ", signif(p_adj_val, 2))
        else                         paste0("p = ", signif(p_adj_val, 2), " (ns)")
      } else { "p = N/A" }
      y_max <- max(alpha_div[[metric]], na.rm = TRUE)
      y_ann <- y_max * 1.08
      ggplot2::ggplot(
        alpha_div,
        ggplot2::aes(
          x = .data[[params$group_col]],
          y = .data[[metric]],
          fill = .data[[params$group_col]],
          text = paste0("Group: ", .data[[params$group_col]],
                        "<br>", label, ": ", signif(.data[[metric]], 4))
        )
      ) +
        ggplot2::geom_boxplot(width = 0.4, outlier.shape = NA, alpha = 0.8, color = "black") +
        ggplot2::geom_jitter(width = 0.12, size = 1.2, alpha = 0.5, show.legend = FALSE) +
        ggplot2::annotate("segment",
          x = 1, xend = 2, y = y_ann, yend = y_ann, color = "black") +
        ggplot2::annotate("text",
          x = 1.5, y = y_ann * 1.02, label = p_label, size = 3.2) +
        ggplot2::scale_fill_manual(values = group_colors) +
        ggplot2::labs(title = label, x = "", y = label) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::theme(legend.position = "none")
    }
    p_list_plotly <- Map(plot_metric_plotly, names(metrics), unname(metrics))
    plotly::subplot(
      lapply(p_list_plotly, function(p) plotly::ggplotly(p, tooltip = "text")),
      nrows = 2,
      margin = 0.04,
      titleX = TRUE,
      titleY = TRUE
    )
  } else if (requireNamespace("patchwork", quietly = TRUE)) {
    p_list[[1]] + p_list[[2]] + p_list[[3]] + p_list[[4]] + patchwork::plot_layout(ncol = 2)
  } else if (requireNamespace("gridExtra", quietly = TRUE)) {
    gridExtra::grid.arrange(grobs = p_list, ncol = 2)
  } else {
    box_tag("plotly, patchwork, or gridExtra is required to combine the four alpha metrics.", "warning")
  }
}
```

### Wilcoxon Test

```{r}
if (is.null(alpha_wil) || !nrow(alpha_wil)) {
  box_tag("The adjusted Wilcoxon table was not found.", "warning")
} else {
  build_dt(alpha_wil[, intersect(c("metric", "comparison", "p_adj"), colnames(alpha_wil)), drop = FALSE], caption = "Wilcoxon test")
}
```

Beta Diversity {data-navmenu="Diversity"}
=====================================

Column {data-width=280}
-------------------------------------

### Beta Diversity

**Fig. 6.** PCoA ordinations based on two beta diversity metrics:

- **Bray-Curtis** — abundance-weighted compositional dissimilarity
- **Jaccard** — presence/absence compositional dissimilarity

Each point is a sample; ellipses show 95% confidence intervals per group.
PERMANOVA R² = fraction of variance explained by group (999 permutations).
Betadisper tests homogeneity of dispersion.

*Weighted and Unweighted UniFrac are computed using the most abundant
representative ASV sequence per species, aligned with MAFFT, and a
neighbor-joining tree. They require `ape` and `GUniFrac` packages and
`taxonomy_filtered.tsv` from the filter step.*

Column {data-width=720, .tabset}
-------------------------------------

```{r beta-pcoa-setup, include=FALSE}
.beta_status <- NULL
.p_bc <- .p_jc <- .p_wuf <- .p_uwuf <- NULL

make_ord <- function(dist_obj, metric_key, title) {
  ord <- cmdscale(dist_obj, k = 2, eig = TRUE)
  eig <- ord$eig; eig[eig < 0] <- 0
  pct <- round(100 * eig / sum(eig), 1)
  d <- data.frame(
    sample_id = rownames(ord$points),
    PC1       = ord$points[, 1],
    PC2       = ord$points[, 2],
    group     = grp,
    stringsAsFactors = FALSE
  )
  perm_row <- if (!is.null(permanova) && "metric" %in% colnames(permanova))
    permanova[permanova$metric == metric_key, , drop = FALSE] else NULL
  ann <- if (!is.null(perm_row) && nrow(perm_row))
    paste0("R2 = ", signif(perm_row$R2[1], 3), " | p = ", signif(perm_row$p_value[1], 3))
  else "PERMANOVA not available"
  p <- ggplot2::ggplot(d, ggplot2::aes(x = PC1, y = PC2, color = group,
            text = paste0("Sample: ", sample_id, "<br>Group: ", group))) +
    ggplot2::geom_point(size = 2.2, alpha = 0.85)
  if (length(unique(d$group)) > 1 && all(table(d$group) >= 3))
    p <- p + ggplot2::stat_ellipse(level = 0.95, linetype = 2)
  p + ggplot2::annotate("text", x = -Inf, y = Inf, label = ann,
          hjust = -0.05, vjust = 1.2, size = 3) +
    ggplot2::scale_color_manual(values = group_colors) +
    ggplot2::labs(title = title,
      x = paste0("PCoA1 (", pct[1], "%)"),
      y = paste0("PCoA2 (", pct[2], "%)"),
      color = "Group") +
    ggplot2::theme_minimal(base_size = 10)
}

make_ord_from_tsv <- function(tsv_df, metric_key, title) {
  if (is.null(tsv_df) || !nrow(tsv_df)) return(NULL)
  req_cols <- c("sample_id", "PC1", "PC2", "group", "pct1", "pct2")
  if (!all(req_cols %in% colnames(tsv_df))) return(NULL)
  tsv_df <- tsv_df[tsv_df$group %in% group_levels, , drop = FALSE]
  if (!nrow(tsv_df)) return(NULL)
  perm_row <- if (!is.null(permanova) && "metric" %in% colnames(permanova))
    permanova[permanova$metric == metric_key, , drop = FALSE] else NULL
  ann <- if (!is.null(perm_row) && nrow(perm_row))
    paste0("R2 = ", signif(perm_row$R2[1], 3), " | p = ", signif(perm_row$p_value[1], 3))
  else "PERMANOVA not available"
  p <- ggplot2::ggplot(tsv_df, ggplot2::aes(x = PC1, y = PC2, color = group,
            text = paste0("Sample: ", sample_id, "<br>Group: ", group))) +
    ggplot2::geom_point(size = 2.2, alpha = 0.85)
  if (length(unique(tsv_df$group)) > 1 && all(table(tsv_df$group) >= 3))
    p <- p + ggplot2::stat_ellipse(level = 0.95, linetype = 2)
  p + ggplot2::annotate("text", x = -Inf, y = Inf, label = ann,
          hjust = -0.05, vjust = 1.2, size = 3) +
    ggplot2::scale_color_manual(values = group_colors) +
    ggplot2::labs(title = title,
      x = paste0("PCoA1 (", tsv_df$pct1[1], "%)"),
      y = paste0("PCoA2 (", tsv_df$pct2[1], "%)"),
      color = "Group") +
    ggplot2::theme_minimal(base_size = 10)
}

if (!has_pkgs(c("ggplot2")) || is.null(tss) || is.null(meta) || !nrow(tss) || !nrow(meta)) {
  .beta_status <- box_tag("There are not enough data to calculate beta diversity ordination.", "warning")
} else if (!requireNamespace("vegan", quietly = TRUE)) {
  .beta_status <- box_tag("The vegan package is not available. Only the PERMANOVA and betadisper tables can be shown.", "warning")
} else {
  meta_beta <- meta[, unique(c(params$sample_id_col, params$group_col)), drop = FALSE]
  beta_df   <- merge(tss, meta_beta, by.x = "sample_id", by.y = params$sample_id_col)
  beta_df   <- beta_df[beta_df[[params$group_col]] %in% group_levels, , drop = FALSE]
  if (nrow(beta_df) < 3L) {
    .beta_status <- box_tag("There are fewer than three filtered samples available to calculate PCoA.", "warning")
  } else {
    mat <- as.matrix(beta_df[, setdiff(colnames(tss), "sample_id"), drop = FALSE])
    rownames(mat) <- beta_df$sample_id
    grp <- factor(beta_df[[params$group_col]], levels = group_levels)
    .p_bc   <- make_ord(vegan::vegdist(mat, method = "bray"), "bray_curtis", "PCoA Bray-Curtis")
    .p_jc   <- make_ord(vegan::vegdist(mat > 0, method = "jaccard", binary = TRUE), "jaccard", "PCoA Jaccard")
    .p_wuf  <- make_ord_from_tsv(pcoa_wunifrac,  "weighted_unifrac",   "PCoA Weighted UniFrac")
    .p_uwuf <- make_ord_from_tsv(pcoa_uwunifrac, "unweighted_unifrac", "PCoA Unweighted UniFrac")
  }
}
```

### Fig. 6a. PCoA — Bray-Curtis

```{r}
if (!is.null(.beta_status)) {
  .beta_status
} else if (isTRUE(pkg_status[["plotly"]])) {
  plotly::ggplotly(.p_bc, tooltip = "text")
} else {
  box_tag("Plotly is not available.", "info")
}
```

### Fig. 6b. PCoA — Jaccard

```{r}
if (!is.null(.beta_status)) {
  .beta_status
} else if (isTRUE(pkg_status[["plotly"]])) {
  plotly::ggplotly(.p_jc, tooltip = "text")
} else {
  box_tag("Plotly is not available.", "info")
}
```

### Fig. 6c. PCoA — Weighted UniFrac

```{r}
if (!is.null(.beta_status)) {
  .beta_status
} else if (is.null(.p_wuf)) {
  box_tag("Weighted UniFrac not available: rerun step 08_beta.R to compute it (requires ape, GUniFrac, and taxonomy_filtered.tsv).", "info")
} else if (isTRUE(pkg_status[["plotly"]])) {
  plotly::ggplotly(.p_wuf, tooltip = "text")
} else {
  box_tag("Plotly is not available.", "info")
}
```

### Fig. 6d. PCoA — Unweighted UniFrac

```{r}
if (!is.null(.beta_status)) {
  .beta_status
} else if (is.null(.p_uwuf)) {
  box_tag("Unweighted UniFrac not available: rerun step 08_beta.R to compute it (requires ape, GUniFrac, and taxonomy_filtered.tsv).", "info")
} else if (isTRUE(pkg_status[["plotly"]])) {
  plotly::ggplotly(.p_uwuf, tooltip = "text")
} else {
  box_tag("Plotly is not available.", "info")
}
```

### Fig. 6e. Beta Diversity Statistics

```{r}
if (!requireNamespace("htmltools", quietly = TRUE)) {
  box_tag("htmltools is not available to arrange both tables side by side.", "warning")
} else {
  htmltools::tagList(
    htmltools::div(
      style = "width:49%;display:inline-block;vertical-align:top;",
      htmltools::tags$h4("PERMANOVA"),
      build_dt(permanova)
    ),
    htmltools::div(
      style = "width:49%;display:inline-block;vertical-align:top;margin-left:1%;",
      htmltools::tags$h4("betadisper"),
      build_dt(betadisper)
    )
  )
}
```

Prevalence and Group-Specific Species {data-navmenu="Composition"}
=====================================

Column {data-width=280}
-------------------------------------

### Prevalence and Group-Specific Species

**Fig. 4a.** Bar and dot chart for species detected exclusively in
`r params$case_label`. Bars = prevalence (left axis); dots = mean relative
abundance (right axis).

**Fig. 4b.** Same for species detected exclusively in `r params$ctrl_label`.

Column {data-width=720}
-------------------------------------

### Fig. 4a. Species Exclusive to `r params$case_label`

```{r excl-case}
req <- c("ggplot2")
if (!has_pkgs(req) || is.null(prev_tbl) || !nrow(prev_tbl) || is.null(case_unique) || !nrow(case_unique)) {
  box_tag("No case-exclusive species are available for this visualization.", "warning")
} else {
  species_col <- colnames(case_unique)[1]
  case_species <- unique(case_unique[[species_col]])
  sp_data <- prev_tbl[prev_tbl$taxon %in% case_species, , drop = FALSE]
  if (!nrow(sp_data)) {
    box_tag("The exclusive-species list does not match the prevalence table.", "warning")
  } else {
    case_prev <- sp_data[sp_data[[params$group_col]] == params$case_label, c("taxon", "prevalence"), drop = FALSE]
    ord <- case_prev$taxon[order(case_prev$prevalence, decreasing = FALSE)]
    sp_data$taxon <- factor(sp_data$taxon, levels = ord)
    max_prev_val   <- max(sp_data$prevalence, na.rm = TRUE)
    max_abund_val  <- max(sp_data$mean_rel_abundance * 100, na.rm = TRUE)
    ab_scale       <- if (is.finite(max_abund_val) && max_abund_val > 0)
                        max_prev_val / max_abund_val else 1
    p_case <- ggplot2::ggplot(sp_data, ggplot2::aes(x = taxon)) +
      ggplot2::geom_bar(
        ggplot2::aes(
          y = prevalence,
          fill = .data[[params$group_col]],
          text = paste0("Taxon: ", taxon, "<br>Prevalence: ", prevalence)
        ),
        stat = "identity",
        position = "dodge",
        alpha = 0.55,
        width = 0.7
      ) +
      ggplot2::geom_point(
        ggplot2::aes(
          y = mean_rel_abundance * 100 * ab_scale,
          color = .data[[params$group_col]],
          text = paste0("Taxon: ", taxon, "<br>Mean relative abundance (%): ", signif(mean_rel_abundance * 100, 3))
        ),
        position = ggplot2::position_dodge(width = 0.7),
        size = 2.8
      ) +
      ggplot2::scale_fill_manual(values = group_colors) +
      ggplot2::scale_color_manual(values = group_colors) +
      ggplot2::scale_y_continuous(
        name     = "Prevalence (N samples)",
        sec.axis = ggplot2::sec_axis(~ . / ab_scale, name = "Mean relative abundance (%)")
      ) +
      ggplot2::coord_flip() +
      ggplot2::labs(x = "", fill = NULL, color = NULL) +
      ggplot2::theme_minimal(base_size = 10)
    if (isTRUE(pkg_status[["plotly"]])) {
      plotly::ggplotly(p_case, tooltip = "text")
    } else {
      p_case
    }
  }
}
```

### Fig. 4b. Species Exclusive to `r params$ctrl_label`

```{r excl-ctrl}
req <- c("ggplot2")
if (!has_pkgs(req) || is.null(prev_tbl) || !nrow(prev_tbl) || is.null(ctrl_unique) || !nrow(ctrl_unique)) {
  box_tag("No control-exclusive species are available for this visualization.", "warning")
} else {
  species_col  <- colnames(ctrl_unique)[1]
  ctrl_species <- unique(ctrl_unique[[species_col]])
  sp_data_c    <- prev_tbl[prev_tbl$taxon %in% ctrl_species, , drop = FALSE]
  if (!nrow(sp_data_c)) {
    box_tag("The control-exclusive species list does not match the prevalence table.", "warning")
  } else {
    ctrl_prev_c  <- sp_data_c[sp_data_c[[params$group_col]] == params$ctrl_label, c("taxon", "prevalence"), drop = FALSE]
    ord_c        <- ctrl_prev_c$taxon[order(ctrl_prev_c$prevalence, decreasing = FALSE)]
    sp_data_c$taxon <- factor(sp_data_c$taxon, levels = ord_c)

    max_prev_c  <- max(sp_data_c$prevalence, na.rm = TRUE)
    max_abund_c <- max(sp_data_c$mean_rel_abundance * 100, na.rm = TRUE)
    ab_scale_c  <- if (is.finite(max_abund_c) && max_abund_c > 0) max_prev_c / max_abund_c else 1

    p_ctrl <- ggplot2::ggplot(sp_data_c, ggplot2::aes(x = taxon)) +
      ggplot2::geom_bar(
        ggplot2::aes(
          y    = prevalence,
          fill = .data[[params$group_col]],
          text = paste0("Taxon: ", taxon, "<br>Prevalence: ", prevalence)
        ),
        stat = "identity",
        position = "dodge", alpha = 0.55, width = 0.7
      ) +
      ggplot2::geom_point(
        ggplot2::aes(
          y     = mean_rel_abundance * 100 * ab_scale_c,
          color = .data[[params$group_col]],
          text  = paste0("Taxon: ", taxon, "<br>Mean relative abundance (%): ",
                         signif(mean_rel_abundance * 100, 3))
        ),
        position = ggplot2::position_dodge(width = 0.7), size = 2.8
      ) +
      ggplot2::scale_fill_manual(values = group_colors) +
      ggplot2::scale_color_manual(values = group_colors) +
      ggplot2::scale_y_continuous(
        name     = "Prevalence (N samples)",
        sec.axis = ggplot2::sec_axis(~ . / ab_scale_c, name = "Mean relative abundance (%)")
      ) +
      ggplot2::coord_flip() +
      ggplot2::labs(x = "", fill = NULL, color = NULL) +
      ggplot2::theme_minimal(base_size = 10)

    if (isTRUE(pkg_status[["plotly"]])) plotly::ggplotly(p_ctrl, tooltip = "text") else p_ctrl
  }
}
```

Differential Abundance {data-navmenu="Informative Species"}
=====================================

Column {data-width=280}
-------------------------------------

### Differential Abundance

**Fig. 7.** Volcano plot of differential abundance results. X-axis: log₂ fold-change
(`r params$case_label` vs `r params$ctrl_label`); Y-axis: −log₁₀(adjusted p-value).
Point size reflects mean relative abundance. Dashed lines mark significance (padj < 0.05)
and effect-size (|log₂FC| ≥ 1) thresholds.

Column {data-width=720}
-------------------------------------

### Volcano plot

```{r}
req <- c("ggplot2")
if (!has_pkgs(req) || is.null(da_res) || !nrow(da_res)) {
  box_tag("No differential abundance results are available.", "warning")
} else {
  da_res$neglog10_padj <- -log10(da_res$padj + 1e-300)
  da_res$point_size <- if ("baseMean_rel" %in% colnames(da_res)) da_res$baseMean_rel else 3
  da_res$sig_chr <- as.character(as.logical(da_res$sig))
  label_df <- da_res[!is.na(da_res$padj) & da_res$padj < 0.05 & abs(da_res$log2fc) >= 1, , drop = FALSE]
  p_volc <- ggplot2::ggplot(
    da_res,
    ggplot2::aes(
      x = log2fc,
      y = neglog10_padj,
      color = sig_chr,
      size = point_size,
      text = paste0(
        "Species: ", feature_id,
        "<br>log2FC: ", signif(log2fc, 4),
        "<br>padj: ", signif(padj, 4)
      )
    )
  ) +
    ggplot2::geom_point(alpha = 0.75) +
    ggplot2::geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
    ggplot2::geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    ggplot2::scale_color_manual(
      values = c("TRUE" = unname(group_colors[params$case_label]), "FALSE" = "grey60"),
      labels = c("TRUE" = "Yes", "FALSE" = "No")
    ) +
    ggplot2::labs(x = "log2 fold-change", y = "-log10(adjusted p-value)", color = "Significant", size = "Mean relative abundance") +
    ggplot2::theme_minimal(base_size = 10)
  if (requireNamespace("ggrepel", quietly = TRUE) && nrow(label_df)) {
    p_volc <- p_volc + ggrepel::geom_text_repel(data = label_df, ggplot2::aes(label = feature_id), size = 2.8, show.legend = FALSE)
  }
  if (isTRUE(pkg_status[["plotly"]])) {
    plotly::ggplotly(p_volc, tooltip = "text")
  } else {
    p_volc
  }
}
```

### Significant Species

```{r}
if (is.null(da_res) || !nrow(da_res)) {
  box_tag("No differential abundance table is available.", "warning")
} else {
  sig_df <- da_res[as.logical(da_res$sig) %in% TRUE, intersect(c("feature_id", "log2fc", "padj", "sig"), colnames(da_res)), drop = FALSE]
  if (nrow(sig_df)) sig_df <- sig_df[order(sig_df$padj), , drop = FALSE]
  build_dt(sig_df)
}
```

Informative Species {data-navmenu="Informative Species"}
=====================================

Column {data-width=280}
-------------------------------------

### Informative Species

Species prioritized by Random Forest (RF) and Linear Discriminant Analysis (LDA).
RF accuracy (CV): `r flags_val("rf_cv_accuracy", "11_informative", "N/A")`

**Fig. 8.** RF feature importance (Mean Decrease Accuracy). Higher values indicate
greater contribution to group discrimination.

**Fig. 9.** LDA coefficients (points) and mean relative abundance by group (bars)
for the top informative species. Positive LD1 = enriched in `r params$case_label`;
negative = enriched in `r params$ctrl_label`. The secondary axis (top) shows the
mean relative abundance scale (%).

**Fig. 10.** PCoA (Bray-Curtis) based only on the top RF informative species.
Restricting the feature space to the most discriminating taxa often reveals
clearer group separation than community-wide ordination.

Column {data-width=720}
-------------------------------------

### Fig. 8. Random Forest Feature Importance

```{r rf-plot}
req <- c("ggplot2")
if (!has_pkgs(req) || is.null(rf_imp) || !nrow(rf_imp)) {
  box_tag("No Random Forest results are available.", "warning")
} else {
  rf_top       <- rf_imp[order(rf_imp$importance, decreasing = TRUE), , drop = FALSE]
  rf_top       <- head(rf_top, 20L)
  rf_top$label <- factor(rf_top$feature_id, levels = rev(rf_top$feature_id))

  p_rf <- ggplot2::ggplot(
    rf_top,
    ggplot2::aes(
      x    = importance,
      y    = label,
      text = paste0("Species: ", feature_id, "<br>Importance: ", signif(importance, 4))
    )
  ) +
    ggplot2::geom_point(size = 2.8, color = "steelblue") +
    ggplot2::geom_segment(
      ggplot2::aes(x = 0, xend = importance, y = label, yend = label),
      color = "grey70", linewidth = 0.5
    ) +
    ggplot2::labs(
      title = "Top 20 informative species (RF)",
      x     = "Mean Decrease Accuracy",
      y     = ""
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(legend.position = "none")

  if (isTRUE(pkg_status[["plotly"]])) plotly::ggplotly(p_rf, tooltip = "text") else p_rf
}
```

### Fig. 9. LDA Coefficients and Mean Relative Abundance

```{r lda-plot}
if (is.null(lda_scores) || !nrow(lda_scores) ||
    is.null(tss) || is.null(meta) ||
    !params$sample_id_col %in% colnames(meta) ||
    !params$group_col %in% colnames(meta)) {
  box_tag("No LDA results are available.", "warning")
} else {
  lda_df      <- lda_scores[order(abs(lda_scores$LD1), decreasing = TRUE), , drop = FALSE]
  lda_df      <- head(lda_df, min(20L, nrow(lda_df)))
  meta_lda    <- meta[, unique(c(params$sample_id_col, params$group_col)), drop = FALSE]
  merged_tss  <- merge(tss, meta_lda, by.x = "sample_id", by.y = params$sample_id_col)
  merged_tss  <- merged_tss[merged_tss[[params$group_col]] %in% group_levels, , drop = FALSE]

  if (!nrow(merged_tss) || !all(lda_df$feature_id %in% colnames(merged_tss))) {
    box_tag("The LDA species do not match the abundance table.", "warning")
  } else {
    mat      <- as.matrix(merged_tss[, lda_df$feature_id, drop = FALSE])
    grp      <- merged_tss[[params$group_col]]
    rs       <- rowSums(mat); rs[rs == 0] <- 1
    mat_tss  <- sweep(mat, 1, rs, "/")
    case_idx <- grp == params$case_label
    ctrl_idx <- grp == params$ctrl_label
    mean_case <- colMeans(mat_tss[case_idx, , drop = FALSE], na.rm = TRUE)
    mean_ctrl <- colMeans(mat_tss[ctrl_idx, , drop = FALSE], na.rm = TRUE)
    total     <- mean_case + mean_ctrl; total[total == 0] <- 1
    plot_df   <- rbind(
      data.frame(feature_id = names(mean_case), side = params$case_label,
                 mean_norm_abundance = (mean_case / total) * 100,  stringsAsFactors = FALSE),
      data.frame(feature_id = names(mean_ctrl), side = params$ctrl_label,
                 mean_norm_abundance = -(mean_ctrl / total) * 100, stringsAsFactors = FALSE)
    )
    plot_df$LD1      <- lda_df$LD1[match(plot_df$feature_id, lda_df$feature_id)]
    ld1_max          <- max(abs(plot_df$LD1), na.rm = TRUE)
    plot_df$LD1_plot <- if (is.finite(ld1_max) && ld1_max > 0) (plot_df$LD1 / ld1_max) * 100 else 0
    plot_df$group    <- ifelse(plot_df$LD1 >= 0, params$case_label, params$ctrl_label)

    feat_order   <- lda_df$feature_id[order(lda_df$LD1)]
    plot_df$label <- factor(plot_df$feature_id, levels = feat_order)

    p_lda <- ggplot2::ggplot(plot_df) +
      ggplot2::geom_col(
        ggplot2::aes(
          x    = label,
          y    = mean_norm_abundance,
          fill = side,
          text = paste0("Species: ", feature_id,
                        "<br>Mean relative abundance: ", signif(abs(mean_norm_abundance), 4), "%")
        ),
        alpha = 0.5
      ) +
      ggplot2::geom_point(
        ggplot2::aes(
          x     = label,
          y     = LD1_plot,
          color = group,
          text  = paste0("Species: ", feature_id, "<br>LD1 (normalized): ", signif(LD1, 4))
        ),
        size = 2.5
      ) +
      ggplot2::geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
      ggplot2::coord_flip() +
      ggplot2::scale_fill_manual(values = group_colors) +
      ggplot2::scale_color_manual(values = group_colors) +
      ggplot2::scale_y_continuous(
        name     = "Normalized LD1 coefficient",
        labels   = function(x) paste0(abs(x)),
        limits   = c(-105, 105),
        sec.axis = ggplot2::sec_axis(
          ~ .,
          name   = "Mean relative abundance (%)",
          labels = function(x) paste0(abs(x))
        )
      ) +
      ggplot2::labs(x = "", fill = NULL, color = NULL) +
      ggplot2::theme_minimal(base_size = 10)

    print(p_lda)
  }
}
```

### Fig. 10. PCoA — Informative Species

```{r infobeta-plot}
if (is.null(info_pcoa) || !nrow(info_pcoa)) {
  box_tag("No informative-species PCoA data available (run step 11 first).", "info")
} else if (!requireNamespace("ggplot2", quietly = TRUE)) {
  box_tag("ggplot2 is required for this plot.", "warning")
} else {
  req_cols <- c("sample_id", "PC1", "PC2", "group", "pct1", "pct2",
                "permanova_R2", "permanova_p")
  if (!all(req_cols %in% colnames(info_pcoa))) {
    box_tag("informative_pcoa.tsv is missing required columns.", "warning")
  } else {
    df_ib <- info_pcoa[info_pcoa$group %in% group_levels, , drop = FALSE]
    if (!nrow(df_ib)) {
      box_tag("No informative-species PCoA rows remain after group filtering.", "info")
    } else {
      ann_ib <- paste0("R2 = ", signif(df_ib$permanova_R2[1], 3),
                       " | p = ", signif(df_ib$permanova_p[1], 3))
      p_ib <- ggplot2::ggplot(
        df_ib,
        ggplot2::aes(x = PC1, y = PC2, color = group,
                     text = paste0("Sample: ", sample_id, "<br>Group: ", group))
      ) +
        ggplot2::geom_point(size = 2.2, alpha = 0.85)
      if (length(unique(df_ib$group)) > 1 && all(table(df_ib$group) >= 3))
        p_ib <- p_ib + ggplot2::stat_ellipse(level = 0.95, linetype = 2)
      p_ib <- p_ib +
        ggplot2::annotate("text", x = -Inf, y = Inf, label = ann_ib,
                 hjust = -0.05, vjust = 1.2, size = 3) +
        ggplot2::scale_color_manual(values = group_colors) +
        ggplot2::labs(
          title = "PCoA Bray-Curtis — informative species only",
          x     = paste0("PCoA1 (", df_ib$pct1[1], "%)"),
          y     = paste0("PCoA2 (", df_ib$pct2[1], "%)"),
          color = "Group"
        ) +
        ggplot2::theme_minimal(base_size = 10)
      if (isTRUE(pkg_status[["plotly"]])) plotly::ggplotly(p_ib, tooltip = "text") else p_ib
    }
  }
}
```
